mosaic benchmark manuscript figures

### 2025-04-10 analysis/manuscript_figures.Rmd
figure clean-up and resizing

```{r currated_VAF, message = FALSE}
(curated_variants_hist <- curated_variants %>%
    mutate(combined_VAF = x_ci_no_Dup / n_ci_no_Dup) %>%
    ggplot(aes(
        x = combined_VAF
    )) +
    geom_vline(
        xintercept = c(0.05, 0.3),
        linetype = "dashed", color = "black"
    ) +
    geom_histogram(
        color = "black",
        position = "identity", bins = 40, na.rm = TRUE
    ) +
    labs(
        x = "Variant Allele Frequency (VAF)",
        y = "# of Variants"
    ) +
    theme_bw() +
    theme(
        strip.text = ggtext::element_markdown()
    ) +
    facet_wrap(~RM_CURATION,
        ncol = 1,
        labeller = labeller(RM_CURATION = c(
            "MOSAIC" = "**Included** in Benchmark",
            "NOT MOSAIC" = "**Excluded** from Benchmark"
        ))
    )
)
```

```{r}
ggsave(
    plot = curated_variants_hist,
    filename = here("figures", "fig3_curated-variants-VAF.pdf"),
    height = 6, width = 4
)

ggsave(
    plot = curated_variants_hist,
    filename = here("figures", "fig3_curated-variants-VAF.png"),
    height = 6, width = 4
)
```

@@

### Paired Plot
```{r}
# Merge direction info into your VAF comparison data
source_vaf_comp <- benchmark_source_comp_df %>%
    filter(
        str_detect(platform, "element|pacbio"),
        !str_detect(platform, "long|sequel")
    ) %>%
    mutate(tech = str_remove(platform, "_.*")) %>%
    group_by(dna_source, tech, CHROM, POS, REF, ALT) %>%
    summarise(VAF = sum(alt_count) / sum(depth), .groups = "drop") %>%
    mutate(
        idx = paste0(CHROM, ":", POS, ":", REF, ":", ALT),
        dna_source = factor(dna_source, levels = c("nonRM", "RM")),
        tech = factor(tech,
            levels = c("element", "pacbio"),
            labels = c("Element", "PacBio Revio")
        )
    ) %>%
    left_join(
        variant_tests %>%
            select(variant_id, direction),
        by = c("idx" = "variant_id")
    ) %>%
    mutate(
        line_color = case_when(
            direction == "Higher in RM" ~ "red",
            direction == "Higher in nonRM" ~ "blue",
            TRUE ~ "lightgrey"
        )
    )

# Generate ggpaired plot with custom line coloring
(source_vaf_plot <- ggpaired(
    source_vaf_comp,
    x = "dna_source", y = "VAF", id = "idx",
    facet.by = "tech",
    line.color = "direction", # use mapped line color
    line.size = 0.4
) +
    ylim(0, 0.4) +
    theme_minimal() +
    theme(legend.position = "none") +
    labs(
        title = "VAF Comparison by DNA Source and Platform",
        subtitle = "Line color indicates VAF direction for significant variants (FDR < 0.05)"
    )
)
```

```{r}
ggsave(
    plot = source_vaf_plot,
    filename = here("figures", "fig4_sample-source-vaf.pdf")
)

ggsave(
    plot = source_vaf_plot,
    filename = here("figures", "fig4_sample-source-vaf.png")
)
```

>>>>

```{r currated_VAF, message = FALSE}
(curated_variants_hist <- curated_variants %>%
    mutate(
        combined_VAF = x_ci_no_Dup / n_ci_no_Dup,
        Variant_Category = case_when(
            RM_KEEP_REMOVE == "KEEP" ~ "**Included** in Benchmark",
            RM_KEEP_REMOVE != "KEEP" ~ "**Excluded** from Benchmark",
            TRUE ~ "ERROR"
        )
    ) %>%
    ggplot(aes(
        x = combined_VAF
    )) +
    geom_vline(
        xintercept = c(0.05, 0.3),
        linetype = "dashed", color = "black"
    ) +
    geom_histogram(
        color = "black",
        position = "identity", bins = 40, na.rm = TRUE
    ) +
    labs(
        x = "Variant Allele Frequency (VAF)",
        y = "# of Variants"
    ) +
    theme_bw() +
    theme(
        strip.text = ggtext::element_markdown()
    ) +
    facet_wrap(~Variant_Category, ncol = 1)
)
```

```{r}
ggsave(
    plot = curated_variants_hist,
    filename = here("figures", "fig3_curated-variants-VAF.pdf"),
    height = 4, width = 4
)

ggsave(
    plot = curated_variants_hist,
    filename = here("figures", "fig3_curated-variants-VAF.png"),
    height = 4, width = 4
)
```

@@

Median benchmark variant coverage for Element short insert and PacBio Revio data.

```{r}
benchmark_source_comp_df %>%
    filter(
        str_detect(platform, "element|pacbio"),
        !str_detect(platform, "long|sequel")
    ) %>%
    mutate(tech = str_remove(platform, "_.*")) %>%
    group_by(dna_source, tech, CHROM, POS, REF, ALT) %>%
    reframe(depth = sum(depth)) %>%
    group_by(dna_source, tech) %>%
    reframe(coverage = median(depth))
```

### Paired Plot
```{r message = FALSE, warning = FALSE}
## custom ggpaired for configurable line.alpha and shape aes
source(here("scripts", "custom_ggpaired.R"))
```

```{r message = FALSE}
# Merge direction info into your VAF comparison data
source_vaf_comp <- benchmark_source_comp_df %>%
    filter(
        str_detect(platform, "element|pacbio"),
        !str_detect(platform, "long|sequel")
    ) %>%
    mutate(tech = str_remove(platform, "_.*")) %>%
    group_by(dna_source, tech, CHROM, POS, REF, ALT) %>%
    summarise(VAF = sum(alt_count) / sum(depth), .groups = "drop") %>%
    mutate(
        idx = paste0(CHROM, ":", POS, ":", REF, ":", ALT),
        dna_source = factor(dna_source, levels = c("nonRM", "RM")),
        tech = factor(tech,
            levels = c("element", "pacbio"),
            labels = c("Element", "PacBio Revio")
        )
    ) %>%
    left_join(
        variant_tests %>%
            select(variant_id, direction, p_adj),
        by = c("idx" = "variant_id")
    ) %>%
    mutate(
        line_color = case_when(
            p_adj > 0.05 ~ "grey20",
            direction == "Higher in RM" ~ "red",
            direction == "Higher in nonRM" ~ "darkblue",
            TRUE ~ "green"
        ),
        line_alpha = case_when(
            p_adj > 0.05 ~ 0.75,
            direction == "Higher in RM" ~ 1,
            direction == "Higher in nonRM" ~ 1,
            TRUE ~ 1
        ),
        dna_source = if_else(dna_source == "nonRM",
            "non-RM", dna_source
        )
    )

# Generate ggpaired plot with custom line coloring
(source_vaf_plot <- ggpaired(
    source_vaf_comp,
    x = "dna_source", y = "VAF", id = "idx",
    facet.by = "tech",
    point.shape = 21,
    point.fill = "line_color",
    point.alpha = "line_alpha",
    line.color = "line_color",
    line.alpha = "line_alpha" # use mapped line color
) +
    ylim(0, 0.4) +
    scale_color_identity() +
    scale_fill_identity() +
    theme_bw() +
    theme(legend.position = "none") +
    labs(
        x = "Sample Source",
        y = "VAF"
    )
)
```

```{r}
ggsave(
    plot = source_vaf_plot, height = 3, width = 6,
    filename = here("figures", "fig4_sample-source-vaf.pdf")
)

ggsave(
    plot = source_vaf_plot, height = 3, width = 6,
    filename = here("figures", "fig4_sample-source-vaf.png")
)
```
